Historical trends in the contribution of six major causes of death to the black-white

### 2017-04-25 Manuscript/Manuscript_COD_Trends.Rmd

---
title: Historical trends in the contribution of six major causes of death to the black-white
  life expectancy gap by US state
bibliography: bibliography.bib
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [ref]. We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf].

@@

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

@@

```{r print-plot-7, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Male.plot[[3]])
```

@@

```{r print-plot-8, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Female.plot[[3]]) 
```

@@

```{r print-plot-9, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Male.plot[[3]]) 
```

@@

```{r print-plot-10, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Female.plot[[3]]) 
```

@@

```{r print-plot-11, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Male.plot`[[3]])
```

@@

```{r print-plot-12, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Female.plot`[[3]])
```

###Discussion 
- the next is to investigate potential state level policy and impact on underlying risk factors
- our analysis highlights a few specific causes in a few specific period
-injuries his CVD, not so much on cancer - because there is more homo. across contribution of cancer
- limitations - major groupings but they have lots of heterogeneity of causes within the grouping
- cont of HIV increases very dramatically, but the pace of decline differed - the places where there is a lag in the decline was the more rural places

@@

- In this paper, we examined state-level trends in six major CODs' contributions to the black-white LE gap.

- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.

- How do these results compare with states with large black populations: Georgia, Florida, Texas, and California? In all of these states, deaths during the first year of life are relatively higher in blacks and in most cases, deaths from CVD during mid-life is also higher among blacks. California and Florida black young men are dying more of injuries, and black men and women in Florida are dying more of communicable disease.

- These findings can be contrasted with those of Firebaugh et al (2014). These authors distinguish between inequality caused by a difference in CVD mortality incidence vs. that caused by a difference in age-at-death and find that while blacks and whites are equally likely to die of heart disease, blacks are dying at younger ages. [qu: how can the equally likely to die of Heart disease part be true when we have the age std mort rates showing otherwise? [*See page 27 of firebaugh - highlighted section - in dropbbox papers directory*].

*Notes of Firebaugh 2014 paper:*
-CVD still contributes the most to the current LE gap. Firebaugh et al (2014) say that policy needs to focus on treatment once CVD is realized, not prevention ("The incidence component is negligible, indicating that blacks and whites are equally likely to succumb to heart disease. Heart disease nevertheless account for one quarter of the gap in 2015 because black victims tends to be younger than white victims. Interventions should focus on identifying heart disease and equalizing treatment. Same goes for Cancers, non-comm diseases. Contrast with HIV/AIDS, prevention is key (since diff is in "incidence", not age at death with comm disease)"
[*Does this reasoning make sense to you? I'm not sure I follow why a diff in age at death would imply focusing on treatment rather than prevention -- what if blacks get CVD at younger ages than whites? Doesn't focusing on equalizing treatment assume that blacks and whites have the same timing of onset?*]

*Literature review*

BW disparities (period-based studies)
- "higher death rates due to heart disease, cancer, homicide (men), diabetes (women), and perinatal conditions" - kochanek et al 2013, harper et al 2007; see also Orsi 2010
- harper 2012; 
- Macinko and Elo (2009): decline in disparity among working-aged people since late 1980s.

@@

We used data over 45 years, across two race/ethnicities, and for all age groups. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs. low *because they had decreased over time*. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ]. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles. [Methodologically, using the Bayesian autoregressive model, allowed us to estimate mortality rates in strata that were sparsely populated...*something about methods?*]

@@

Secondly, our analysis does not account for immigration, or migration between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014]. As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect life expectancy. Migration is particularly important for the District of Columbia, because of it's small geographical bounds (it is easier to be considered migration when regions are smaller - how to say this?), and substantial in-migration from New York, in- and out-migration to contiguous Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. These changes would both serve to increase the racial gap in DC.

helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
-movement of southern blacks to northern states, and migration out of rust belt.
-DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge out flux to MD and Virginia. Should almost consider these three places as one.
-lots of migration into MD
-NY - 24% are immigrants
-Texas has influx of immigrants to 17%
-lots of migration into Virginia, as well as immigration
-Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
-Alabama gets migrants from other southern states
-Connecticut - influx in those immigrating from outside US since 1990s

Individuals who migrate are more likely to have completed a bachelor's degree than those who do not, suggesting that migration may contribute to heightened life expectancy.

Thus, states with black or white populations that have shifted over time will be affected. This limits the interpretation of results especially for Washington DC, as it has undergone substantial demographic shifts since X. 
-foreign borns tend to have higher LE, so increase in immigration can lead to gains in LE in either blacks or whites, or both, depending on the country of origin - better because of migrant selectivity (select on traits related to good health) and better health (less likely to smoke or be obese or binge drink [ref], and return migration when they become sick)

Problems specific to DC
-migration's effect on LE over the course of the time period

-migration - has been on the decline since the 1980s (Molloy 2011, 2017). migration happens more in people with bachelor's degrees in their early twenties (and declines thereafter) but is at a bit of a higher rate throughout lifetime than those without degree (Kaplan 2017)
-roughly 1.5% of the popn moves between census regions per year, and 1.3% move within states in a census region.

Read: racial disparity in where people from DC move:
-https://ggwash.org/view/38326/theres-a-big-racial-disparity-in-where-people-born-in-dc-end-up-moving-or-not-moving

*Conclusions*

- DC exhibits the highest contributions for many causes, but they have the greatest gap and a widening gap - so more room for each cause to contribute. Goes to say that all causes are contributing to the gap (it isn't being driven by one cause in particular)

Life Expectancy generally
The spotlight on racial disparities in the US has intensified. Several recent papers examining US mortality by region, cause of death, or within population subgroups have received considerable attention while ignoring important black-white disparities.2, 3, 4 These disparities remain profound, and no previous paper has examined cause-specific contributions by state over an extended time period. Our analysis provides an important contribution to the literature by highlighting a high degree of heterogeneity in state-specific performance over time in eliminating these inequalities.
from above:
2. Chetty R, Stepner M, Abraham S, Lin S, Scuderi B, Turner N, Bergeron A, Cutler D. The association between income and life expectancy in the United States, 2001-2014. JAMA. 2016. 315(16):1750-1766.

3. Dwyer-Lindgren L, Bertozzi-Villa A, Stubbs RW, Morozoff C, Kutz MJ, Huynh C, Barber RM, et al. US County-Level Trends in Mortality Rates for Major Causes of Death, 1980-2014. JAMA. 2016. 316(22):2385-2401.

4. Case A, Deaton A. Rising morbidity and mortality in midlife among white non-Hispanic Americans in the 21st century. Proc Natl Acad Sci. 2015. 112(49):15078-83.

Cancer

Oliver. CEBP. 2008. Recent trends in BW disparities in Cancer Mortality. Note: higher mort rates in their paper likely because they restrict to >35yo.
-BW diff in lung cancer mort mimics trends in overall mort rate in males. In females, no diff in lung cancer (although minor difference in other cancer related to smoking)
-increasing disparity is seen in colorectal cancer in males between B and W
-hump also seen in prostate cancer, especially among blacks, but falling mort rates since mid 1990s.
-among females, increasing disparity in mort rates from colorectal cancer and breast cancer since 1975 (beginning of their study)
-not sure what explains the inverted-U in cancer mort in our data among females - this CEBP people doesn't seem to shed light on this.

Screening - when is it proven effective vs. when does it affect the recorded COD? Corinne to review that literature.
-breast cancer: kills mostly among those 70+ yo.

-trends in incidence may be related to overdx (prostate cancer, for ex) - for prostate cancer don't know whether the decline is due to improvements in treatment or due to screening -- could be both or just one

https://nccd.cdc.gov/uscs/toptencancers.aspx#White
-black women have higher rates of breast cancer than white women (28/100000 vs 20 for mortality)
-blacks have higher rates of prostate cancer than whites (39.1 vs 18)

Smoking and lung cancer
see: https://www.washingtonpost.com/news/to-your-health/wp/2015/11/12/smoking-among-u-s-adults-has-fallen-to-historic-lows-these-7-charts-show-who-still-lights-up-the-most/?utm_term=.f2b261e63acf -- uses CDC data for citation
-smoking rates have fallen steadily over the course of the study time period, but homogeneity in smoking still exists. used to be 42.4% in 1965 and down to 16.8% in 2014.
-by race: nearly same for blacks and whites - 17.5% in blacks vs. 18.2% in whites, but 11.2% in hispanics (which are mixed into BW designation in our study)
-by geography: highest in the midwest (20.7%), then south (17%), NE (15.3%), then west (13%), within heterogeneity within state (
Dwyer-Lindgren  L, Mokdad  AH, Srebotnjak  T, Flaxman  AD, Hansen  GM, Murray  CJ.  Cigarette smoking prevalence in US counties: 1996-2012.  Popul Health Metr. 2014;12(1):5.)

-Delancey et al (2008) identified more rapid decreases in smoking related cancers among black males since 1990 as contributing to the decrease in the overall gap.

Obesity and...
-high obesity in the south - ref 9
-women and blacks have highest rates of obesity - ref 8

Opiods
https://www.cdc.gov/nchs/products/databriefs/db273.htm - Although rates increased 63% for non-Hispanic black persons (from 7.5 per 100,000 in 1999 to 12.2 in 2015) and 43% for Hispanic persons (from 5.4 in 1999 to 7.7 in 2015), non-Hispanic white persons experienced the greatest increase between 1999 and 2015. The age-adjusted rate of drug overdose deaths among non-Hispanic white persons increased 240% from 6.2 per 100,000 in 1999 to 21.1 in 2015.

Centers for Disease Control and Prevention. Vital signs: overdoses of prescription opioid pain relievers and other drugs among women—United States, 1999–2010. MMWR Morb Mortal Wkly Rep. 2013;62(26):537–542: opioid-related mortality increased 415% among women and 265% among men between 1999 and 2010.38

Paulozzi LJ, Xi Y. Recent changes in drug poisoning mortality in the United States by urban-rural status and by drug type. Pharmacoepidemiol Drug Saf. 2008;17(10):997–1005.: in 1999, large central metropolitan areas had the highest opioid-related mortality rates, and noncore areas had the lowest rates; by 2004, noncore areas had the highest rates and had seen the largest relative increase during that time period.50

Paulozzi LJ, Ryan GW. Opioid analgesics and rates of fatal drug poisoning in the United States. Am J Prev Med. 2006;31(6):506–511: drug poisoning mortality rates varied 8-fold between states, as did opioid consumption (methadone, 13-fold; oxycodone, 7-fold; overall, 4-fold).17

Centers for Disease Control and Prevention. Vital signs: overdoses of prescription opioid pain relievers—United States, 1999–2008. MMWR Morb Mortal Wkly Rep. 2011;60(43):1487–1492.: Another study showed that prescription opioid sales ranged from 3.7 (Illinois) to 12.6 (Florida) kilograms per 10,000 population, and opioid-related mortality rates ranged from 5.5 (Nebraska) to 27 (New Mexico) per 100,000 population.2

-should the focus be prevention of specific condictions/diseases, treatment of those conditions/diseases, or more distal --- SES differences - edu, employment, income (Hummer Du Bois review argues we need to focus on SES distal exposures, while Firebaugh 2014 differentiates whhen to focus on treatmnet of condition vs. prevention of the condition)

-medicaid coverage varies among states. MS and NY have longer black LE than general trend and also have expanded medicare coverage (Bharmal 2012 citing the henry kaiser family foundation). CA NJ in middle for LE and coverage, AL TX lower for both -- so there is this observed ecological association

@@

>>>>

---
title: Historical trends in the contribution of six major causes of death to the black-white
  life expectancy gap by US state
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks [ref nchs_LE]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013[ref: seer stat]. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System [ref: nchs data]. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

For the analysis we excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010 [ref Census_map]. We did not consider Hispanic origin to identify race according to Hispanic ethnicity, since the quality of the coding of ethnicity varied by state over the time period included [ref cdc_hisp].

@@

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [@selvin_text; @preston_demography_text]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and sex.

@@

```{r print-plot-7, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Male.plot[[3]])

`Injuries.Male.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) + 
           geom_hline(aes(yintercept = 6), lwd = 0.5, col = "grey", alpha = 0.5)

@@

```{r print-plot-8, fig.width = 9, fig.height = 9, echo = F}
#ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Injuries.Female.plot[[3]])

`Injuries.Female.plot`[[3]] +
           geom_hline(aes(yintercept = -0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
           geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5)
```

@@

```{r print-plot-9, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Male.plot[[3]])

`Communicable.Male.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) 
```

@@

```{r print-plot-10, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(Communicable.Female.plot[[3]])

`Communicable.Female.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

@@

```{r print-plot-11, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Male.plot`[[3]])

`All other causes.Male.plot`[[3]] +
  geom_hline(aes(yintercept = -1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 3), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5) 
```

@@

```{r print-plot-12, fig.width = 9, fig.height=9, echo = F}
#ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
#ggplotly(`All other causes.Female.plot`[[3]])

`All other causes.Female.plot`[[3]] +
  geom_hline(aes(yintercept = -2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 4), lwd = 0.5, col = "grey", alpha = 0.5)
```

###Discussion

@@

In this paper, we examined state-level trends in six major CODs' contributions to the black-white life expectancy gap. While we found that many states mimicked the national patterns of cause-of-death specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. We found that cardiovascular disease contributes the most to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries continue to play a role in contributing to the current gap for males in several states, especially in the Midwest and DC, while contributing to a decreased gap among females in the South. Illinois, Michigan and Wisconsin also show relatively high contributions from cancer. On the other hand, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contribution of more than two years. Thus, the reason for New York’s and Massachusetts’s low CVD contribution does not appear to be the effect of recent changes, but highlights a long-standing minimal difference and point to the importance of viewing the most recent estimates in light of their earlier values.

**Literature review**

This research extends previous studies that have estimated either the contribution of causes of death to the life expectancy gap on the national level, or the excess risk among blacks in cause of death specific mortality rates (references: Harper 2007, 2012; kochanek 2013, Orsi 2010).

@@

We used data over 45 years, across two race/ethnicities, and for all age groups. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs. low *because they had decreased over time*. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ]. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles.

@@

Secondly, our analysis does not account for immigration, or migration between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014]. As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect the estimated life expectancy of a state. Migration is particularly important for the District of Columbia, because of it's small geographical bounds, and substantial in-migration from New York, in- and out-migration to Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Such changes would both serve to increase the racial gap in DC.

Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time.

This analysis provides a foundation for future work to investigate time-varying and time-fixed state-level characteristics that may be responsible for differences among state in risk factors leading to each cause of death. Differences in smoking policy, medicare coverage, access to care, segregation, and racism, for example, could all affect risk factors for ill health and vary in magnitude by state.

@@

to add to bibliography list:
nchs_LE ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf

seer stat: National Cancer Institute. Surveillance, Epidemiology, and End Results
Program: SEER*Stat software: latest release: version 8.1.5 [Internet]. Bethesda (MD): NCI; 2013 Mar 31 [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/seerstat

nchs data: National Cancer Institute. Surveillance, Epidemiology, and End Results Program: US mortality data [Internet]. Bethesda (MD): NCI; [cited 2014 Jun 20]. Available from: http://www.seer.cancer.gov/mortality/

cdc_hisp: [https://www.cdc.gov/nchs/data/series/sr_02/sr02_148.pdf]

census_map: https://www.census.gov/2010census/popmap/. retrived apr 25, 2017
